Automatic Differentiation Accelerates Inference for Partially Observed Stochastic Processes

### 2024-05-10 paper/ms.tex
Updates from Overleaf

@@

As mentioned earlier, the correction factors accumulate over time. If $\theta$ and $\phi$ coincide, one can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$, as seen in Equation \ref{eq:weighting-scheme}, while still targeting the posterior and likelihood at $\theta$. This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate. When $\alpha=1$, MOP-$1$ maintains the memory of each particle's ancestral trajectory, recovering the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}. When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{blei2018vsmc}. $\alpha$ therefore controls an exponentially-weighted moving average, determining how much "memory" the resulting gradient estimate has over ancestral trajectories. This is illustrated in Figure \ref{fig:biasvar}, in the context of the Dhaka cholera model of \cite{king08}.

@@

@@

We show here that MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood. While the result is presented here as specific to MOP-$\alpha$, the full proof relies on a handful of technical lemmas presented in the supplementary material that may be useful in their own right, and Theorem \ref{thm:mop-targeting} can also serve as a more general result as a strong law of large numbers for triangular arrays of particles with ``off-policy'' resampling.

@@

@@

\cite{scibior21} show that the estimate of \cite{blei2018vsmc} is the gradient of a vanilla particle filter when resampling terms are dropped, and also recover the estimate of \cite{poyiadjis11} through applying a``stop-gradient trick" to the particle filter. It turns out that both of these, when applied on the bootstrap filter, correspond to special cases of MOP-$\alpha$.

@@

\subsection{MOP-$\alpha$ Bias and Variance}
\begin{thm}
    \label{thm:mop-biasvar}
\end{thm}

\subsection{Convergence Rates}
\begin{thm}
    \label{thm:mop-convergence}
\end{thm}

>>>>

\kevin{Can someone who's reading through the words and glazing through the math get the idea that (1) we're running AD on the PF, and (2) using an off-policy resampling trick to get our estimator. Should I hedge by writing the SI in Del Moral language? }

@@

As mentioned earlier, the correction factors accumulate over time. If $\theta$ and $\phi$ coincide, one can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$, as seen in Equation \ref{eq:weighting-scheme}, while still targeting the posterior and likelihood at $\theta$. This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate. When $\alpha=1$, MOP-$1$ maintains the memory of each particle's ancestral trajectory, recovering the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}. When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{blei2018vsmc}.

\cite{scibior21} show that the estimate of \cite{blei2018vsmc} is the gradient of a vanilla particle filter when resampling terms are dropped, and also recover the estimate of \cite{poyiadjis11} through applying a ``stop-gradient trick" to the particle filter. It turns out that both of these, when applied on the bootstrap filter, correspond to special cases of MOP-$\alpha$.

$\alpha$ therefore controls an exponentially-weighted moving average, determining how much "memory" the resulting gradient estimate has over ancestral trajectories. This is illustrated in Figure \ref{fig:biasvar}, in the context of the Dhaka cholera model of \cite{king08}.

@@

\subsection{Linear Convergence Rates}
\begin{thm}
    \label{thm:mop-convergence}
\end{thm}

@@

We show here that MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood. While the result is presented here as specific to MOP-$\alpha$, we actually prove a more general result in the supplementary material. That is, we show a strong law of large numbers for triangular arrays of particles with ``off-policy'' resampling, where we resample the particles according to an arbitrary resampling rule not necessarily in proportion to the targeted distribution of interest, to weights that encode the cumulative discrepancy between the resampling and the target distribution instead of equal weights.

\kevin{Should we just show the general result? That is, for an arbitrary set of indices $k_{n,j}$ generated in proportion to some arbitrary weights $\pi_{n,j}$ that do not necessarily sum to $1$ as the estimate is self-normalized, reweighting the particles by $w_{n,a(j)}g_{n,a(j)}/\pi_{n,a(j)}$ targets the posterior. $\pi_{n,j}$ can depend on $\{(x_{n,1:J},g_{n,1:J})\}$ as long as the resampling is carried out independently of $\{(x_{n,1:J},g_{n,1:J})\}$, conditional on $\pi_{n,1:J}$}

@@

\kevin{Write the supplement formally with Del Moral-lite language, going into full detail with the proofs and calculations.}

@@

@@

\subsection{MOP-$\alpha$ Bias and Variance}
\begin{thm}
    \label{thm:mop-biasvar}
    When $\alpha\in(0,1)$, $\theta=\phi$, the MSE and variance of MOP-$\alpha$, where $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)(1-\alpha)^{-1}$, are:
    \vspace*{-1ex}
    \begin{align*}
        &\E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 
        \\
        &\lesssim \min_{k \leq N} NpG'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi(\alpha)\right),\\[-1ex]
        &\text{Var}(\nabla_\theta \hat\ell^{\alpha}(\theta)) \lesssim \min_{k\leq N} NpG'(\theta)^2\left(\frac{k^2}{(1-\alpha)^2J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N\right).
        \end{align*}
\end{thm}
\kevin{Explain exactly what $\epsilon$ is here, and interpret the rates as agreeing with our intuition. Talk about the relationship between the theory and the empirical results in the bias-variance tradeoff figure above, as well as the phase transition between $Nk^2/J$ and $N^4/J$ variance with MOP-$\alpha$ and MOP-$1$, where $k$ is really a surrogate for $\alpha$ -- with more memory, you truncate less. }

We defer the proof to the supplementary material in favor of a very brief outline here.

\kevin{Say that our framework suggests whole new classes of algorithms, as it includes Naesseth and Poyiadjis. Interpolating between them with discounting is only one way to interpolate, we can also e.g. truncate at a fixed lag. Remark that the analysis is similar for MOP-$(1,k)$ as a fixed-lag filter, with slightly better rates but a less natural interpretation. We leave empirical studies of its performance to future work. }

### 2024-05-10 presentation/presentationmath.tex
Updates from Overleaf

\documentclass{beamer}
\setbeamertemplate{caption}[numbered]
\usepackage{tikz}
\usepackage{graphicx}
\usetheme{AnnArbor}
%\usecolortheme{wolverine}
\title[AD for POMPs]{Automatic Differentiation Accelerates Inference for Partially Observed Stochastic Processes}
\author[Kevin Tan]{Kevin Tan \\ \small{University of Pennsylvania, Department of Statistics and Data Science} \\\vspace{0.5cm} \small{Joint work with Giles Hooker and Edward Ionides}}

@@

    \begin{itemize}
        \item Unobserved Markov process $\{X_t, t \geq t_0\}$, observations $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$. Write $X_n := X_{t_n}$.
        \item \pause Unknown parameter $\theta \in \Theta$, to be estimated.
        \item $X_t \in \gX \subseteq \R^d$, $Y_n \in \gY$, $\theta \in \Theta \subseteq \R^p$.\pause
        \item \textbf{Notation:}
        \begin{itemize}
            \item $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ is the \textbf{process model}.
            \item \pause $\texttt{process}\left(x_n, \theta\right)$ is the \textbf{simulator} corresponding to the process model. 
            \item \pause $f_{Y_n|X_n}\left(y_n \mid x_n, \theta\right)$ is the \textbf{measurement model}.
        \end{itemize}
    \end{itemize}
\end{frame}

@@

    \begin{itemize}
        \item Unobserved Markov process $\{X_t, t \geq t_0\}$, observations $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$.
        \item \pause Problems to solve:
        \begin{itemize}
            \item \pause The \textbf{filtering problem}:
            \begin{itemize}
                \item \pause Find the density of the hidden state at time $n$, given all observations until time $n$.
                \item \pause $f_{X_n|Y_{1:n}}(x_n|y_{1:n}^*;\theta)$.
            \end{itemize}
            \item \pause Related problem, estimating the \textbf{posterior} distribution of hidden states:
            \begin{itemize}
                \item \pause Find the density of the hidden states at timesteps $1$ to $n$, given all observations till time $n$. 
                \item \pause $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*;\theta)$.
            \end{itemize}
        \end{itemize}
    \end{itemize}
\end{frame}

@@

    \begin{table}[h!]
        \centering
        \begin{tabular}{||c c||} 
         \hline
         Authors & Method \\ [0.5ex] 
         \hline\hline
         \cite{poyiadjis11} & Particle approximation of score
         \\ & $O(N^2)$ variance, unbiased \\
         \hline
         \cite{blei2018vsmc} & Backprop through vanilla PF \\ 
         & Asymptotic bias \\
         \hline
         \cite{corenflos21} & Optimal transport resampling \\
         & Consistent, $O(J^2)$ runtime \\
         \hline
         \cite{scibior21} & Stop-gradient trick \\ 
         & Recovers \cite{poyiadjis11} with AD \\
         \hline 
         \cite{singh22} & Fixed-lag smoothing \\ 
         & Need transition densities \\ 
         \hline
        \end{tabular}
        \label{table:lit-review}
    \end{table}
\end{frame}

@@

\begin{frame}{Bias-Variance Tradeoff with $\alpha$}
\begin{itemize}
    \item $\alpha$ balances a tradeoff between:
    \begin{itemize}
        \item \pause Maintaining memory of each particle's ancestral trajectory (most extreme when $\alpha=1$).
        \item \pause Considering only the single-step transition dynamics (when $\alpha=0$).
    \end{itemize}
    \item \pause\textbf{Exponentially-weighted moving average} where $\alpha$ controls the amount of discounting.
    \item \pause Bias small if $y_{n+1:N}^*$ not informative for $x_n$ given $y_{0:n}^*$ (\cite{corenflos21}).
\end{itemize}

@@

\begin{frame}{Particle MCMC}
    \begin{itemize}
        \item One can use the 
    \end{itemize}
\end{frame}

>>>>

\documentclass{beamer}
\setbeamertemplate{caption}[numbered]
\usepackage{tikz}
\usepackage{graphicx}
\usetheme{AnnArbor}
%\usecolortheme{wolverine}
\title[AD for POMPs]{Automatic Differentiation Accelerates Inference for Partially Observed Stochastic Processes}
\author[Kevin Tan]{Kevin Tan \\ \small{University of Pennsylvania, Department of Statistics and Data Science\\\vspace{0.5cm} \small{Advised by Giles Hooker and Edward Ionides}}}

@@

    \begin{itemize}
        \item Unobserved Markov process $\{X_t, t \geq t_0\}$, observations $Y_1,...,Y_N \in \gY$ taking values $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$. Write $X_n := X_{t_n} \subseteq \R^d$.
        \item \pause Unknown parameter $\theta \in \Theta \subseteq \R^p$, to be estimated.
        \item \textbf{Notation:}
        \begin{itemize}
            \item $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ is the \textbf{process model}.
            \item \pause $\texttt{process}\left(x_n, \theta\right)$ is the \textbf{simulator} corresponding to the process model. 
            \item \pause $f_{Y_n|X_n}\left(y_n \mid x_n, \theta\right)$ is the \textbf{measurement model}.
        \end{itemize}
    \end{itemize}
\end{frame}

@@

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=1.5]

\node[black] (X_0) at (-3,0) {$X_0$};
\node[black] (Xdot0) at (-2,0) {$...$};
\node[black] (X_{n-1}) at (-1,0) {$X_{n-1}$};
\node[black] (X_n) at (0,0) {$X_n$};
\node[black] (X_{n+1}) at (1,0) {$X_{t+1}$};
\node[black] (Xdot{n+1}) at (2,0) {$...$};
\node[black] (X_N) at (3,0) {$X_T$};
\node[black] (Y_0) at (-3,-1) {$Y_0$};
\node[black] (Ydot0) at (-2,-1) {$...$};
\node[black] (Y_{n-1}) at (-1,-1) {$Y_{n-1}$};
\node[black] (Y_n) at (-0,-1) {$Y_n$};
\node[black] (Y_{n+1}) at (1,-1) {$Y_{n+1}$};
\node[black] (Ydot{n+1}) at (2,-1) {$...$};
\node[black] (Y_N) at (3,-1) {$Y_N$};
\path[->, black] (X_0) edge (Xdot0);
\path[->, black] (Xdot0) edge (X_{n-1});
\path[->, black] (X_{n-1}) edge (X_n);
\path[->, black] (X_n) edge (X_{n+1});
\path[->, black] (X_{n+1}) edge (Xdot{n+1});
\path[->, black] (Xdot{n+1}) edge (X_N);
\path[->, black] (X_0) edge (Y_0);
\path[->, black] (X_{n-1}) edge (Y_{n-1});
\path[->, black] (X_n) edge (Y_n);
\path[->, black] (X_{n+1}) edge (Y_{n+1});
\path[->, black] (X_N) edge (Y_N);
\end{tikzpicture}
\caption{Directed acyclic graph (DAG) of a POMP model. The latent states $(X_t)_{t_0\leq t \leq T}$ follow a Markov process, and the observations $Y_n$ are conditionally independent given the latent state. }
\label{fig:pomp}
\end{figure}
\end{frame}

    \begin{itemize}
        \item Problems to solve:
        \begin{itemize}
            \item \pause The \textbf{filtering problem}:
            \begin{itemize}
                \item \pause Find the density of the hidden state at time $n$, given all observations until time $n$, $f_{X_n|Y_{1:n}}(x_n|y_{1:n}^*;\theta)$.
            \end{itemize}
            \item \pause Related problem, estimating the \textbf{posterior} distribution of hidden states:
            \begin{itemize}
                \item \pause Find the density of the hidden states at timesteps $1$ to $n$, given all observations till time $n$, $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*;\theta)$.
            \end{itemize}
        \end{itemize}
    \end{itemize}
\end{frame}

@@

    \begin{table}[h!]
        \centering
        \begin{tabular}{||c c||} 
         \hline
         Authors & Method \\ [0.5ex] 
         \hline\hline
         \cite{poyiadjis11} & Particle approximation of score.
         \\ & $O(N^2)$ variance, consistent. \\
         \hline
         \cite{blei2018vsmc} & Backprop through vanilla PF. \\ 
         & Asymptotic bias. \\
         \hline
         \cite{corenflos21} & Optimal transport resampling. \\
         & Consistent, $O(J^2)$ runtime. \\
         \hline
         \cite{scibior21} & Stop-gradient trick. \\ 
         & Recovers \cite{poyiadjis11} with AD. \\
         \hline 
         \cite{singh22} & Fixed-lag smoothing. \\ 
         & Need transition densities. \\ 
         \hline
        \end{tabular}
        \label{table:lit-review}
    \end{table}
\end{frame}

@@

\begin{frame}{Bias-Variance Tradeoff with $\alpha$}
\begin{itemize}
    \item $\alpha$ balances a tradeoff between:
    \begin{itemize}
        \item \pause Maintaining memory of each particle's ancestral trajectory (most extreme when $\alpha=1$).
        \item \pause High-variance, but consistent.
        \item \pause Considering only the single-step transition dynamics (when $\alpha=0$).
        \item \pause Low-variance, but asymptotically biased.
    \end{itemize}
    \item \pause\textbf{Exponentially-weighted moving average} where $\alpha$ controls the amount of discounting.
    \item \pause Bias small if $y_{n+1:N}^*$ not informative for $x_n$ given $y_{0:n}^*$ (\cite{corenflos21}).
\end{itemize}

@@

\begin{frame}{Particle MCMC}
    One can use the likelihood estimate from the particle filter to perform MCMC. But the random walk proposal is very bad!

\begin{figure}[t!]
    \centering
    \includegraphics[width = \textwidth/\real{1.5}]{imgs/pmcmc/mh.png}
    \caption{Convergence diagnostics for the Metropolis-Hastings variant particle MCMC with the random walk proposal and an informative empirical prior from IF2. Here, we display the results for the trend parameter in the Dhaka cholera model of \cite{king08}.}
    \label{fig:mh}
\end{figure}
\end{frame}

\begin{frame}{MOP-$\alpha$ Speeds up the Mixing of Particle MCMC}
    Using the gradient estimate from MOP-$\alpha$ within a NUTS sampler (initialized with an informative prior estimated with IF2) works wonders -- burn-in period of $700,000$ in \cite{fasiolo16} reduced to $500$ here.
    \begin{figure}[t!]
    \centering
    \includegraphics[width=\textwidth/\real{1.5}]{imgs/pmcmc/nuts_eb.png}
    \caption{Convergence diagnostics for a No U-Turn Sampler with the informative prior from IF2. The NUTS sampler mixes quickly, and the posterior estimates from each chain posterior modes.}
    \label{fig:nuts-eb}
\end{figure}
